---
title: "Rule Resource Usage and Default Tuning"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

<style>
div.main-container,
.main-container {
  max-width: 1800px !important;
  width: 96% !important;
}

@media (min-width: 992px) {
  .main-container > .row {
    display: flex;
    align-items: flex-start;
  }

  .main-container > .row > div:first-child {
    flex: 0 0 300px;
    max-width: 300px;
    align-self: flex-start;
    position: sticky;
    top: 20px;
    max-height: calc(100vh - 40px);
    overflow-y: auto;
  }

  .main-container > .row > div:last-child {
    flex: 1 1 auto;
    max-width: calc(100% - 300px);
  }

  .toc-content {
    max-width: none !important;
    width: 100% !important;
    padding-left: 24px !important;
    padding-right: 24px !important;
  }

  div.tocify,
  #TOC {
    position: relative !important;
    top: auto !important;
    width: 100% !important;
    max-width: 280px !important;
    max-height: none !important;
    overflow-y: visible !important;
  }
}

pre {
  white-space: pre !important;
  overflow-x: auto !important;
  font-size: 12px;
  line-height: 1.25;
}

table {
  width: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5
)

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(yaml)

repo_root <- normalizePath(file.path(getwd(), "..", ".."))
repo_file <- function(...) file.path(repo_root, ...)
```

## Question

`config-template.yaml`'s `rule_resources` block had 17 rules with hand-set,
never-benchmarked `mem_mb`/`cpus`/`time` values (some as high as 240 GB / 24
hours), alongside a `default` of 4000 MB / 1 cpu / 60 min that ~30 other rules
silently relied on. We added `benchmark:` to every real (non-orphaned) rule in
the pipeline and ran the full workflow-tests hamster dataset end to end twice,
using **every chromosome in the real reference genome** (21 chromosomes across
the `autosomes`/`xy` groups, ~2.72 Gb total) rather than the 3-chromosome
subset used for day-to-day testing - once with `run_phylop: false` (the
original scope of this analysis), then again with `run_phylop: true` to fill
in the `phylop_regions.smk` rules and pick up real data for `maf_index`, which
hadn't re-executed the first time. This notebook uses that combined real
`benchmark:` output to answer: which of the current custom entries are
actually justified, which are wildly over-provisioned, how does usage scale
with chromosome/group size, and where does run time actually concentrate
across the pipeline's stages.

## Data and methodology

```{r load-data}
`%||%` <- function(a, b) if (is.null(a)) b else a

bench <- read_tsv(repo_file("analyses", "resource-usage", "benchmarks-flat.tsv"), show_col_types = FALSE) %>%
  mutate(
    wall_min = wall_s / 60,
    across(c(max_rss_mb, cpu_time_s, mean_load_pct), as.numeric)
  )

fai_path <- "/n/holylfs05/LABS/informatics/Everyone/support/20251114-raftrey-hamster-wga/data/genomes/masked/GCA_000001635.9/GCA_000001635.9.fna.masked.hard.fna.fai"
chrom_lengths <- read_tsv(fai_path, col_names = c("chrom", "length_bp", "offset", "linebases", "linewidth"), show_col_types = FALSE) %>%
  select(chrom, length_bp)

test_cfg <- read_yaml(repo_file("data", "hamsters", "workflow-tests", "workflow-tests.yaml"))
groups <- test_cfg$ref_chromosome_groups
group_totals <- tibble(
  group = names(groups),
  group_length_bp = sapply(groups, function(chrs) sum(chrom_lengths$length_bp[match(unlist(chrs), chrom_lengths$chrom)]))
)

rule_cfg <- read_yaml(repo_file("config-template.yaml"))$rule_resources
default_cfg <- rule_cfg$default

cfg_table <- bind_rows(lapply(names(rule_cfg), function(r) {
  x <- rule_cfg[[r]]
  tibble(rule = r, cfg_mem_mb = x$mem_mb %||% default_cfg$mem_mb,
         cfg_cpus = x$cpus %||% default_cfg$cpus,
         cfg_time_min = x$time %||% default_cfg$time,
         is_custom = r != "default")
}))

bench <- bench %>%
  left_join(chrom_lengths, by = "chrom") %>%
  left_join(group_totals, by = "group") %>%
  mutate(scale_bp = coalesce(length_bp, group_length_bp))
```

- **Data source**: primarily `data/hamsters/workflow-tests/test-full-gc/`, a
  real (non-dry-run), error-free, 361/361-step execution against the real
  mouse reference genome and real 15-species Cactus alignment used throughout
  `workflow-tests.yaml` - all 21 real chromosomes across the `autosomes`/`xy`
  groups, `rho_mode: fixed`, and `sample_file` blank (auto-selecting the new
  MAF-based GC path, `get_gc_content_from_maf`, exercised here for the first
  time - see its own section below). `run_phylop` was off for this run, so
  phyloP-stage rules and `get_gc_content` (the older sample_file-based path)
  are carried over unchanged from the earlier `test-full-phylop-rho/` run
  where they still have data - see [Caveats](#caveats) for what that means for
  those specific rows.
- **`benchmark:`** is Snakemake's built-in per-job resource tracker: it records
  wall-clock time, peak RSS memory, CPU time, and mean CPU load for the
  process tree of each job, written to `logs/benchmarks/<rule>/...txt`.
- 407 individual job records across 25 rules were parsed.
- **Not covered by this run**: `phastcons_chunk` and
  `phastcons_estimate_rho_chunk` (orphaned rules, never scheduled - see repo
  `RULES.md`).

## Run time by pipeline stage

Before the per-rule detail below, a higher-level view: every rule grouped
into the pipeline stage it belongs to (matching the `NN-name` output
directories the workflow itself writes to), showing the real distribution of
wall-clock time across all 21 chromosomes/groups per rule.

```{r stage-map}
stage_map <- c(
  ref_fasta_index = "01: MAF prep", ref_fasta_dict = "01: MAF prep", maf_index = "01: MAF prep",
  make_group_beds = "01: MAF prep", maf_split_chr_by_group = "01: MAF prep",
  picard_scatter_by_ns = "01: MAF prep", ns_to_bed = "01: MAF prep", filter_ns_bed_minlen = "01: MAF prep",
  ns_minlen_to_bed3 = "01: MAF prep", fixed_windows_bed = "01: MAF prep", maf_index_chr = "01: MAF prep",
  maf_split_chunks = "01: MAF prep", filter_maf_by_gap = "01: MAF prep", ref_gff_split_by_chr = "01: MAF prep",

  extract_4d_codons_by_chr = "02: Neutral model", extract_4d_sites = "02: Neutral model",
  filter_4d_sites = "02: Neutral model", run_phylofit = "02: Neutral model",
  get_gc_content = "02: Neutral model", get_gc_content_from_maf = "02: Neutral model",
  run_mod_freqs = "02: Neutral model",

  run_phylop = "03: phyloP", convert_wig_to_bed = "03: phyloP", adjust_pvals = "03: phyloP",
  get_conserved_sites = "03: phyloP", get_accelerated_sites = "03: phyloP",
  site_counts = "03: phyloP", naive_cluster_conserved_sites = "03: phyloP",

  run_phastcons_chr = "04: phastCons", global_rho = "04: phastCons", phastcons_concat_chr = "04: phastCons",

  extract_cds_bed_chr = "05: CNEEs", cnees_from_conserved_chr = "05: CNEEs",
  cnees_to_bed4_chr = "05: CNEEs", cnee_alignments_chr = "05: CNEEs"
)

stage_data <- bench %>%
  mutate(stage = stage_map[rule]) %>%
  filter(!is.na(stage))
```

```{r stage-runtime-plot, fig.height=5.5}
ggplot(stage_data, aes(x = stage, y = wall_min)) +
  geom_boxplot(outlier.shape = NA, fill = NA, color = "grey50") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.5, size = 2, color = "#3b6e73") +
  scale_y_log10(labels = scales::label_number()) +
  labs(x = "Pipeline stage", y = "Run time (minutes, log scale)",
       title = "Run time by pipeline stage, all rules and chromosomes") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor.y = element_blank())
```

Unlike a single Cactus-style pipeline with a handful of coarse steps, this
distribution spans many small bookkeeping rules alongside genuinely heavy
compute, so a log scale is used (the linear scale that fits a 3-step pipeline
cleanly would flatten everything below `phastCons` into an unreadable sliver
here) - each point is one rule-chromosome job, not one rule. Reading it: **01:
MAF prep** and **05: CNEEs** are dominated by fast bookkeeping steps with a
handful of slower outliers (`maf_split_chunks`, `cnee_alignments_chr`); **02:
Neutral model** is bimodal, mixing near-instant BED/summary rules with the
much slower `extract_4d_codons_by_chr`; **03: phyloP** and **04: phastCons**
are the two genuinely compute-heavy stages, with `phastCons` carrying both
the highest median and the widest spread - consistent with it being the only
stage that re-runs an HMM-like scan per chunk rather than a single pass per
chromosome.

```{r stage-mem-plot, fig.height=5.5}
ggplot(stage_data, aes(x = stage, y = max_rss_mb)) +
  geom_boxplot(outlier.shape = NA, fill = NA, color = "grey50") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.5, size = 2, color = "#8a5a44") +
  scale_y_log10(labels = scales::label_number()) +
  labs(x = "Pipeline stage", y = "Peak RSS (MB, log scale)",
       title = "Peak memory by pipeline stage, all rules and chromosomes") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor.y = element_blank())
```

Memory tells a different story than time: every stage's typical (median) job
sits in the low hundreds of MB, and the spread within each stage is far
narrower than the time plot's - the same handful of rules
(`maf_split_chr_by_group`, `extract_4d_codons_by_chr`, `picard_scatter_by_ns`)
account for essentially all of the upper tail across every stage, rather than
whole stages running memory-heavy the way `phastCons`/`phyloP` run
time-heavy.

## Per-rule usage vs. configured resources

```{r summary-table}
summary_tbl <- bench %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    peak_mem_mb = max(max_rss_mb),
    mean_mem_mb = mean(max_rss_mb),
    peak_wall_min = max(wall_min),
    mean_wall_min = mean(wall_min),
    peak_load_pct = max(mean_load_pct),
    .groups = "drop"
  ) %>%
  left_join(cfg_table, by = "rule") %>%
  mutate(
    cfg_mem_mb = coalesce(cfg_mem_mb, default_cfg$mem_mb),
    cfg_cpus = coalesce(cfg_cpus, default_cfg$cpus),
    cfg_time_min = coalesce(cfg_time_min, default_cfg$time),
    is_custom = coalesce(is_custom, FALSE),
    mem_pct_of_cfg = 100 * peak_mem_mb / cfg_mem_mb,
    time_pct_of_cfg = 100 * peak_wall_min / cfg_time_min
  ) %>%
  arrange(desc(mem_pct_of_cfg))

summary_tbl %>%
  select(rule, is_custom, n, peak_mem_mb, cfg_mem_mb, mem_pct_of_cfg,
         peak_wall_min, cfg_time_min, time_pct_of_cfg, cfg_cpus, peak_load_pct) %>%
  mutate(across(where(is.numeric), \(x) round(x, 1))) %>%
  kable(caption = "Peak observed usage vs. currently configured resources, sorted by memory pressure")
```

**Headline finding**: every rule's peak memory usage in this full-scale run is
under the `default` bucket's 4000 MB - including `maf_split_chr_by_group`
(configured 240,000 MB, used 3246 MB: **1.4%**) and `extract_4d_codons_by_chr`
(configured 128,000 MB, used 3626 MB: **2.8%**). None of the current custom
`mem_mb`/`time` values appear to be grounded in measured usage; they read as
defensive guesses.

Three rules do sit close enough to `default`'s ceiling (all in the 78-91%
range) that I would not rely on bare `default` for them without a margin:
`picard_scatter_by_ns`, `maf_split_chr_by_group`, and `extract_4d_codons_by_chr`.
Everything else currently custom (`cnee_alignments_chr`, `phastcons_concat_chr`,
`cnees_from_conserved_chr`, `filter_maf_by_gap`, `maf_index_chr`, `run_phylofit`)
peaked at or below 33% of `default` and has no real need for its own entry,
other than `cnee_alignments_chr`'s `cpus: 4` (see below).

## CPU allocation check

`mem_mb`/`time` aren't the only thing worth checking - `mean_load_pct` (average
percent of one core used over the job's lifetime) indicates whether the
`cpus:` a rule requests is actually being used.

```{r cpu-check}
summary_tbl %>%
  mutate(load_vs_requested = peak_load_pct / (cfg_cpus * 100)) %>%
  select(rule, cfg_cpus, peak_load_pct, load_vs_requested) %>%
  arrange(desc(load_vs_requested)) %>%
  mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
  kable(caption = "Peak mean CPU load vs. requested cpus (>1.0 = using more than requested)")
```

`cnee_alignments_chr` (`cpus: 4`, since it passes `-p <cpus>` to `mafutils
fetch`) shows load consistent with genuine multi-core use, so its `cpus: 4`
looks justified regardless of how much its memory entry shrinks.
`maf_split_chunks` shows load above 100% of its single requested cpu - worth
a look, though not acted on here since it's currently `default`-reliant and
not part of the custom-entry cleanup this notebook is scoped to.

## Scaling with chromosome/group size

For the 21 per-chromosome rules, peak memory and wall time against the actual
chromosome length they operated on, to see whether usage scales with sequence
size (and if so, how steeply) rather than being roughly constant:

```{r scaling-plot, fig.height=9}
per_chrom_rules <- bench %>%
  filter(!is.na(length_bp)) %>%
  distinct(rule) %>%
  pull(rule)

plot_data <- bench %>% filter(rule %in% per_chrom_rules, !is.na(length_bp))

ggplot(plot_data, aes(x = length_bp / 1e6, y = max_rss_mb)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5) +
  facet_wrap(~rule, scales = "free_y") +
  labs(x = "Chromosome length (Mb)", y = "Peak RSS (MB)",
       title = "Peak memory vs. chromosome length, per rule") +
  theme_bw(base_size = 10)
```

```{r scaling-plot-time, fig.height=9}
ggplot(plot_data, aes(x = length_bp / 1e6, y = wall_min)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5) +
  facet_wrap(~rule, scales = "free_y") +
  labs(x = "Chromosome length (Mb)", y = "Wall time (min)",
       title = "Wall time vs. chromosome length, per rule") +
  theme_bw(base_size = 10)
```

```{r scaling-fits}
fit_stats <- plot_data %>%
  group_by(rule) %>%
  filter(n() >= 5) %>%
  summarise(
    mem_slope_mb_per_mb = coef(lm(max_rss_mb ~ I(length_bp / 1e6)))[2],
    mem_r2 = summary(lm(max_rss_mb ~ I(length_bp / 1e6)))$r.squared,
    .groups = "drop"
  ) %>%
  arrange(desc(mem_slope_mb_per_mb))

fit_stats %>%
  mutate(across(where(is.numeric), \(x) round(x, 3))) %>%
  kable(caption = "Linear fit of peak memory (MB) per Mb of chromosome length")
```

Most rules show a shallow or near-flat memory/time slope against chromosome
length - consistent with tools whose cost is dominated by fixed per-job
overhead (parsing, I/O setup) rather than genuinely scaling with sequence
size at this genome's scale. Only three rules combine a real slope with a
fit that actually explains the variance (R² > 0.75): `cnee_alignments_chr`
(6.07 MB/Mb, R²=0.76), `maf_split_chunks` (4.68 MB/Mb, R²=0.84), and
`filter_maf_by_gap` (0.08 MB/Mb, R²=0.86) - all three process whole-chromosome
alignment content directly, which fits. Notably, `extract_4d_codons_by_chr`
has a nonzero slope (2.42 MB/Mb) but essentially no explanatory power
(R²=0.01) - its memory use doesn't meaningfully track chromosome length in
this data, so whatever drives its usage (species count, 4d-site density,
something else) isn't captured by length alone. Projecting the well-fit
rules' slopes to a chromosome twice the length of this genome's largest
(~390 Mb) stays well under 8 GB - reassuring, but this is a single-species,
single-genome-size extrapolation and should not be treated as validated for
meaningfully larger genomes (e.g. many-Gb plant or amphibian
genomes) without re-running this same analysis on such a dataset.

## Group-level rules

`maf_split_chr_by_group` and `make_group_beds` run once per *group*, not per
chromosome, so there are only two real data points (`autosomes`, `xy`) -
too few to fit a trend, but worth showing since `maf_split_chr_by_group`
carries the largest current default (240 GB):

```{r group-level-table}
bench %>%
  filter(rule %in% c("maf_split_chr_by_group", "make_group_beds")) %>%
  mutate(group_length_mb = round(scale_bp / 1e6, 1), max_rss_mb = round(max_rss_mb, 1), wall_min = round(wall_min, 2)) %>%
  select(rule, group, group_length_mb, max_rss_mb, wall_min, mean_load_pct) %>%
  kable(caption = "Group-level rule usage (autosomes = 19 chromosomes / ~2.46 Gb, xy = 2 chromosomes / ~0.26 Gb)")
```

The `autosomes` group (~2.46 Gb, 19 chromosomes processed in one job) used
3246 MB - about 1.3 MB of RSS per Mb of input sequence. Even a substantially
larger genome assembled into one group would need to be roughly 60x this
genome's size before approaching the currently-configured 240 GB ceiling.

## Recommended tiers {#recommendations}

**Update: applied.** The recommendation below was implemented in
`config-template.yaml` (18 keys - `default` + 17 ungrounded custom entries -
down to 13: `default` + 12), organized around *why* a rule's cost varies
rather than raw magnitude:

- **Unaffected by genome size or species count** (pure BED/coordinate
  arithmetic or post-hoc summary, never touches raw alignment content) -
  dropped to plain `default`: `cnees_from_conserved_chr`,
  `convert_wig_to_bed`, `naive_cluster_conserved_sites`,
  `phastcons_concat_chr`, `ref_fasta_dict`.
- **Affected by genome size only** (scans the reference genome, never touches
  the multi-species alignment) - kept custom with a modest margin:
  `picard_scatter_by_ns` (real peak already 78% of `default`'s ceiling on
  this genome).
- **Affected by both genome size and species count** (reads or holds actual
  alignment content, whose width scales with species count) - kept
  individually custom, each flagged in the file as "benchmarked at n=15
  species - re-validate before scaling species count up": `maf_index`,
  `maf_split_chr_by_group`, `extract_4d_codons_by_chr`, `run_phylofit`,
  `run_phylop`, `maf_index_chr`, `maf_split_by_ns_minlen`,
  `filter_maf_by_gap`, `phastcons_chunk`, `cnee_alignments_chr` (the only
  rule needing `cpus > 1`, confirmed by the CPU-load check above).
- **`adjust_pvals`** was left untouched (`mem_mb: 24000`, `cpus: 32`) rather
  than reclassified - its `cpus: 32` looks like intentional parallelism
  (`resources.cpus_per_task` is passed straight into
  `utils/adjust_pvals.sh`), not something to fold into a generic tier without
  understanding it first.

The second test run (with `run_phylop: true`) has since confirmed several of
these were sound: `maf_index` peaked at only 156 MB against its configured
8000 MB (2%) - essentially matching its per-chromosome sibling
`maf_index_chr` - suggesting it could be trimmed further in a future pass.
`run_phylop` peaked at 2458 MB (31% of its 8000 MB), and `adjust_pvals` at
4250 MB (18% of its untouched 24000 MB) - both comfortably within their
current settings. All five rules newly confirmed on plain `default`
(`convert_wig_to_bed`, `get_conserved_sites`, `get_accelerated_sites`,
`site_counts`, `naive_cluster_conserved_sites`) peaked at 131-142 MB (~3.5% of
`default`), validating that they didn't need their own entry.

## global_rho (rho_mode: estimate): from timeout to real numbers

`global_rho` under `rho_mode: estimate` has a different story from every
other rule here - not an ungrounded guess being over- or under-provisioned,
but an actual production incident. Real cluster runs this session showed the
two chromosomes with the most chunks (`CM000994.3`: 138,496 chunks,
`CM000995.3`: 129,961) repeatedly timing out - first at 180 min (the tier it
originally shared with `run_phastcons_chr`), then 480, then 720 - even after
adding `ThreadPoolExecutor`-based concurrency across `cpus_per_task` threads.

`sacct` accounting on both the failed jobs and several successfully-completed
`global_rho` jobs (chunk counts 99k-114k) all showed the same signature: only
~28% CPU efficiency (~2.2-2.3 of the 8 allocated cores actually busy),
regardless of chunk count or success/failure - a systemic ceiling, not
something specific to the two large chromosomes. The cause was filesystem
I/O, not CPU or `phastCons` itself: each chunk's `phastCons --estimate-rho`
call wrote its own transient working files, plus a per-chunk `.rho.txt` and
`.stderr.log`, all to the same shared network-filesystem directory from 8
concurrent threads - a small-file/metadata-heavy pattern that bottlenecks
network filesystems under concurrency. The sibling rule `run_phastcons_chr`,
which threads over the same shape of per-chunk work but writes exactly 3
known files directly to their final path (no globbing, no redundant
sidecar), never showed this problem.

The fix: route `phastCons`'s transient `--estimate-rho` output through
Snakemake's per-job scratch resource (`resources.tmpdir`, typically
node-local) instead of the shared output filesystem, only write
`.stderr.log` for chunks that fail, and drop the redundant per-chunk
`.rho.txt` (the value was already being aggregated in memory) unless
`debug_keep_intermediates` is set. None of this touched the rho estimation
logic itself.

```{r global-rho-table}
bench %>%
  filter(rule == "global_rho") %>%
  mutate(wall_min = round(wall_min, 2), max_rss_mb = round(max_rss_mb, 1)) %>%
  select(group, chrom, wall_min, max_rss_mb) %>%
  arrange(desc(wall_min)) %>%
  kable(caption = "global_rho (rho_mode: estimate) real usage across all 21 chromosomes, post-fix")
```

Post-fix, all 21 chromosomes completed - including the two that had never
finished before - with a real peak of 31.2 minutes (`CM000994.3`) and 831 MB
(`CM001002.3`, notably not the largest chromosome by chunk count; worth
another look if it recurs). Re-checking `sacct` on that same `CM000994.3`
job post-fix showed efficiency jump to ~77% (~6.1 of 8 cores busy),
confirming the diagnosis. Against the interim placeholder values used while
chasing this down (`mem_mb: 8000`, `cpus: 8`, `time: 720`), real usage is
under 11% of the memory budget and under 5% of the time budget - both safe
to bring down substantially (e.g. `mem_mb: 2000-2500`, `time: 90-120`, a
2.5-4x margin over the observed peak), while keeping `cpus: 8` since it's
now delivering genuine parallelism rather than mostly idling.

**Note:** `mean_load_pct` is omitted from the table above deliberately -
Snakemake's own benchmark tracker does not reliably capture CPU time spent
in subprocess children spawned from a `ThreadPoolExecutor` inside a `run:`
block (as `global_rho`, `run_phastcons_chr`, and `cnee_alignments_chr` all
do), so it reads misleadingly low (single digits) even when `sacct`
job-level accounting - which does aggregate the whole process tree - shows
genuine multi-core use. See [Caveats](#caveats).

## get_gc_content_from_maf and summary_report: first real data

Both are new this session and had never executed for real before `test-full-gc` -
`get_gc_content_from_maf` computes genome-wide GC directly from the whole input MAF via
`mafutils gc` (the new default GC source, auto-selected whenever `sample_file` is blank,
replacing the external accession-lookup path); `summary_report` collects the pipeline
summary HTML report described elsewhere in this repo. Both currently fall back to plain
`default` (`mem_mb: 4000`, `cpus: 1`, `time: 60`) - no custom entry existed for either
before this run.

```{r new-rules-table}
bench %>%
  filter(rule %in% c("get_gc_content_from_maf", "summary_report")) %>%
  mutate(wall_min = round(wall_min, 2), max_rss_mb = round(max_rss_mb, 1)) %>%
  select(rule, wall_min, max_rss_mb, mean_load_pct) %>%
  kable(caption = "First real benchmark data for the two newest rules")
```

`get_gc_content_from_maf` took 25.8 minutes and peaked at 134 MB, running essentially
single-core the whole time (`mean_load_pct` ~98%, `cpu_time_s` almost exactly equal to
wall time) - consistent with `mafutils gc`'s own documentation: GC counting itself is
cheap per base, but the sequential path (`-p 1`, the default) is bound by per-base Python
overhead, not CPU work to parallelize. Parallel mode (`-p > 1`) works for uncompressed or
bgzip-compressed MAFs alike (not bgzip-only, correcting an earlier claim in this notebook's
own history) - this pipeline's real MAF is uncompressed, so parallel mode was a real option
worth actually measuring rather than assuming.

**Update: real `-p 4` (`cpus: 4`) data now exists**, from `config-template.yaml` being set to
`cpus: 4` for this rule and exercised for real in `test-full-split-aln`. First attempt, at
the interim `mem_mb: 750` carried over from the sequential-mode recommendation above, OOM
killed. Bumped to `mem_mb: 4000` to unblock, which succeeded - but real usage came in
*above* even that:

```{r gc-parallel-table}
bench_numseqs_gc <- read_tsv(repo_file("analyses", "resource-usage", "benchmarks-flat-num_seqs.tsv"), show_col_types = FALSE) %>%
  filter(rule == "get_gc_content_from_maf") %>%
  mutate(across(c(max_rss_mb, cpu_time_s, mean_load_pct), as.numeric))

bench_numseqs_gc %>%
  mutate(wall_min = round(wall_s / 60, 2), max_rss_mb = round(max_rss_mb, 1)) %>%
  select(wall_min, max_rss_mb, cpu_time_s) %>%
  kable(caption = "get_gc_content_from_maf, cpus: 4 (-p 4), real data from test-full-split-aln")
```

486.19s wall (8m06s) and **5,922 MB peak** - well above the `mem_mb: 4000` it actually ran
under (this run's success is closer to lucky than validated; SLURM/cgroup accounting
sometimes has slack above the nominal request). `cpu_time_s` (349.91) is well under 4x
wall time despite `cpus: 4`, meaning parallel mode isn't using anywhere near 4 full cores
of work here for this MAF - the wall-time win over sequential mode (25.8min -> 8.1min,
~3.2x) is real, but at a real memory cost (134MB -> 5.9GB, ~44x) disproportionate to that
speedup. Recommend a real margin over the observed peak, e.g. `mem_mb: 7000-7500`, rather
than treating the current `4000` as validated just because this one run happened to
succeed.

`summary_report` took well under a minute and 190 MB - comfortably inside `default` with
no changes needed.

## split_strategy: num_seqs vs ns - first real comparison

`split_strategy: num_seqs` is new this session and is now the pipeline's default,
replacing `ns` (splitting on N-runs in the reference FASTA) as the default chunk-splitting
method for phastCons. Instead of the reference genome, it splits on runs of MAF blocks with
few aligned species (`num_seqs`, already recorded in the block index `mafutils index`
produces) - no reference-masking dependency, and it reflects actual cross-species
alignability rather than an indirect proxy for it. See
`analyses/resource-usage/resource-usage-and-defaults.Rmd` git history and this repo's
Claude session notes for the exploratory analysis that motivated it (Cactus tiles the
reference continuously with zero true coverage gaps, so pure "MAF coverage" splitting is a
dead end; `num_seqs` run-length is the real signal).

This section compares its first full real run
(`test-full-split-aln/`, `num_seqs_max_for_gap: 3`, `num_seqs_min_gap_bp: 100`,
`num_seqs_min_keep_region_len: 6`) against the existing `ns`-based `test-full-gc/` run -
same real 21-chromosome dataset, same `rho_mode: fixed`.

### Timing

```{r num-seqs-vs-ns-timing}
bench_numseqs <- read_tsv(repo_file("analyses", "resource-usage", "benchmarks-flat-num_seqs.tsv"), show_col_types = FALSE) %>%
  mutate(across(c(max_rss_mb, cpu_time_s, mean_load_pct), as.numeric))

chunk_driven_rules <- c("maf_split_chunks", "filter_maf_by_gap", "run_phastcons_chr",
                         "phastcons_concat_chr", "cnee_alignments_chr", "cnees_from_conserved_chr",
                         "cnees_to_bed4_chr", "maf_index_chr", "extract_cds_bed_chr")

timing_compare <- bind_rows(
  bench %>% filter(rule %in% chunk_driven_rules) %>% mutate(strategy = "ns"),
  bench_numseqs %>% filter(rule %in% chunk_driven_rules) %>% mutate(strategy = "num_seqs")
) %>%
  group_by(rule, strategy) %>%
  summarise(n = n(), sum_wall_s = round(sum(wall_s), 1), sum_cpu_s = round(sum(cpu_time_s), 1), .groups = "drop") %>%
  pivot_wider(names_from = strategy, values_from = c(n, sum_wall_s, sum_cpu_s))

timing_compare %>%
  kable(caption = "Chunk-count-driven rules, summed across all 21 chromosomes: ns vs num_seqs")
```

Every chunk-count-driven rule got faster, most of them dramatically:
`maf_split_chunks` ~4.2x less wall time, `filter_maf_by_gap` ~5.2x,
`phastcons_concat_chr` ~13x, `run_phastcons_chr` ~1.8x. Summed across every job in the
pipeline, total wall-clock dropped from 135,190s (37.6h) to 79,851s (22.2h) - a ~1.7x
reduction - while total CPU-seconds barely changed (32,514s to 30,520s, ~6% less). That
split (big wall-time win, flat CPU time) is exactly what the earlier per-chunk phastCons
timing test (see git history) predicted: `num_seqs` isn't doing less real computational
work, it's paying far less *overhead* per chunk and far less filesystem contention from
having ~18x fewer files per chromosome - the same mechanism behind the `global_rho`
timeout saga above. `phastcons_concat_chr`'s 13x drop is the clearest single confirmation,
since that rule is pure glob-and-concatenate over one file per chunk.

One number not to over-read: `run_phastcons_chr`'s recorded `cpu_time_s` went up slightly
(1,804s to 2,269s) despite wall time dropping - this is one of the `ThreadPoolExecutor` +
subprocess-children rules already flagged in [Caveats](#caveats) as having unreliable CPU
accounting, so that specific figure shouldn't be read as a real regression.

### Biological output: CEs and CNEEs

Total conserved-element (CE) basepairs are essentially unchanged between strategies, but
`ns`'s far smaller chunks call many more, shorter elements - consistent with chunk
boundaries artificially fragmenting what `num_seqs`'s larger contiguous chunks let
phastCons call as one continuous element instead.

```{r ce-cnee-compare}
tibble(
  metric = c("CEs: count", "CEs: total bp", "CEs: median length (bp)",
             "CNEEs: count", "CNEEs: total bp", "CNEEs: median length (bp)"),
  ns = c(2127094, 140939708, 31, 575157, 86119768, 103),
  num_seqs = c(1554862, 139193307, 49, 592809, 91245906, 107)
) %>%
  kable(format.args = list(big.mark = ","),
        caption = "CE/CNEE totals across all 21 chromosomes, ns vs num_seqs (04-phastcons/regions, 05-cnees bed4)")
```

A direct overlap check on `CM000994.3` (`bedtools jaccard` between the two strategies' CE
beds) confirms this is the same underlying signal, not a different result: **Jaccard =
0.81**, with ~88-90% of each strategy's CE basepairs overlapping the other's.

Final CNEEs (post-CDS-subtraction, length-filtered) are modestly higher under `num_seqs`
(+3% count, +6% bp) - plausibly because it processes more of the genome in the first place
(176.4 Mb vs 112.5 Mb on `CM000994.3` alone; it doesn't exclude real repeat-masked sequence
the way `ns`/hard-masking does), so there's simply more raw material in which a conserved
element could be found.

**Net read:** `num_seqs` is faster on every real metric measured, and finds essentially the
same conservation signal as `ns` with less chunk-boundary fragmentation. `num_seqs_max_for_gap`,
`num_seqs_min_gap_bp`, and `num_seqs_min_keep_region_len` are still an exploratory starting
point (see `config-template.yaml`'s comments) rather than a fully calibrated default - this
was the first real full-pipeline run of the new strategy, not a systematic sweep across
threshold values.

## Caveats {#caveats}

- **Single species, single genome size.** All of this is one ~2.7 Gb mouse
  reference genome with a 15-species alignment. The scaling fits above are
  illustrative, not validated for larger/smaller genomes or different
  alignment depths (more/fewer species).
- **`ref_fasta_index`/`ref_fasta_dict`/`maf_index` (whole-genome) and all
  phyloP-stage rules are missing from this pass again.** They had real data
  earlier this session (see git history of this file), but those outputs
  live outside `test-full-gc`'s own directory (reference-adjacent) or simply
  weren't re-executed (`run_phylop` was off for this run), so they're absent
  from both of this pass's source directories. Nothing wrong with those
  rules' current settings - this is a data-availability gap, not a
  regression - but their `stage-runtime-plot`/`stage-mem-plot` groupings
  will show fewer or no points until a run that re-executes them is folded
  back in. `ref_fasta_dict` is worth remembering specifically: it OOM'd once
  against its `default` `mem_mb: 4000` earlier this session before
  succeeding on retry at 2601 MB peak - real run-to-run variance exists here
  (Picard's `CreateSequenceDictionary`), worth a margin above any single
  measured peak rather than trusting it as a hard ceiling.
- **CPU load is average, not peak-instantaneous** for rules that don't spawn
  subprocess children - `mean_load_pct` is a time-averaged figure from
  Snakemake's benchmark tracker and can understate brief bursts of
  multi-core use.
- **CPU load is unreliable, not just averaged, for rules that spawn
  subprocess children from a `ThreadPoolExecutor`** (`global_rho`,
  `run_phastcons_chr`, `cnee_alignments_chr`). Snakemake's benchmark tracker
  appears to miss most CPU time spent in forked/exec'd child processes
  (e.g. `phastCons` subprocesses), so `mean_load_pct` reads as low single
  digits for these rules even when `sacct` job-level accounting - which
  aggregates the whole process tree via the cgroup - shows genuine
  multi-core use (confirmed directly for `global_rho`: ~7% by
  `mean_load_pct` vs. ~77% by `sacct` on the same job). Use `sacct`, not
  this notebook's `mean_load_pct` column, to judge real CPU efficiency for
  any subprocess-spawning threaded rule.
