---
title: "CE/CNEE Predictions vs. UCSC mm39 35-way Conserved Elements"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

<style>
div.main-container,
.main-container {
  max-width: 1800px !important;
  width: 96% !important;
}

@media (min-width: 992px) {
  .main-container > .row {
    display: flex;
    align-items: flex-start;
  }

  .main-container > .row > div:first-child {
    flex: 0 0 300px;
    max-width: 300px;
    align-self: flex-start;
    position: sticky;
    top: 20px;
    max-height: calc(100vh - 40px);
    overflow-y: auto;
  }

  .main-container > .row > div:last-child {
    flex: 1 1 auto;
    max-width: calc(100% - 300px);
  }

  .toc-content {
    max-width: none !important;
    width: 100% !important;
    padding-left: 24px !important;
    padding-right: 24px !important;
  }

  div.tocify,
  #TOC {
    position: relative !important;
    top: auto !important;
    width: 100% !important;
    max-width: 280px !important;
    max-height: none !important;
    overflow-y: visible !important;
  }
}

pre {
  white-space: pre !important;
  overflow-x: auto !important;
  font-size: 12px;
  line-height: 1.25;
}

table {
  width: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5
)

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

repo_root <- normalizePath(file.path(getwd(), "..", ".."))
repo_file <- function(...) file.path(repo_root, ...)
```

## Question

Our phastCons-derived CEs (`04-phastcons/regions`) and CNEEs
(`05-cnees/phastcons/bed/*.cnees.bed4`) come from a 15-species Cactus alignment. As an
independent sanity check, how well do they agree with UCSC's `mm39-35way.bed` - conserved
elements called from a completely different, much larger (35-species) multiz alignment,
using a completely different pipeline (phastCons on multiz, not Cactus)? Agreement here
isn't proof of correctness, but strong disagreement would be a red flag; strong agreement
is a meaningful, independent corroboration of what our pipeline is finding. This also
compares our two `split_strategy` runs (`num_seqs`, now the default, and `ns`, from
[[resource-usage-and-defaults.Rmd]]'s comparison) against this same external reference.

## Methodology

Our reference (`GCA_000001635.9`) is the exact same assembly as UCSC's mm39/GRCm39 -
confirmed directly from the FASTA header ("GRCm39 reference primary assembly"), the
reference GFF's filename, and NCBI's own `sequence_report.tsv` for this assembly. Same
assembly means same coordinates - no liftover needed, only a chromosome-name translation
(UCSC's `chr1` etc. to GenBank accessions like `CM000994.3`), via
`translate_ucsc_chroms.sh` in this directory (reusing the header-lookup `awk` idiom already
used for exactly this in the support data repo's `04_replace_bed_seqids_with_genbank.sh`),
restricted to the 21 real chromosomes this pipeline actually has predictions for.

Per chromosome, per run (`num_seqs`, `ns`), per our-track (CE, CNEE), `compare_to_ucsc.py`
runs `bedtools jaccard` and `bedtools intersect` (via `bedtools-env`'s `bedtools`, called by
absolute path rather than `conda activate` - this session hit a real bug twice where
`conda activate` doesn't reliably win on `$PATH`) between our bed and the translated UCSC
bed for that chromosome, producing `ucsc-comparison-flat.tsv`.

**A real pipeline quirk surfaced and worked around while building this**: the raw CE bed
(`04-phastcons/regions`) uses phastCons's own truncated chromosome name (e.g. `CM000994`,
no version suffix) rather than the full accession every other output uses - the same thing
`cnees_from_conserved_chr` already has to normalize internally when reading this same file
to build CNEEs. `compare_to_ucsc.py` does the same normalization before comparing. Relatedly,
the on-disk `04-phastcons/regions/{group}/{chrom}.bed` *file* sits next to a same-named
`{chrom}/` *directory* (holding `run_phastcons_chr`'s small `chunks.done`/`global_rho.txt`
bookkeeping files, left behind after `cleanup_chunk_intermediates` removes the real per-chunk
outputs) - not a bug, but an easy thing to overlook browsing the output tree.

For the CNEE track specifically, UCSC's track has no CDS subtracted while ours does, so the
raw comparison undersells the real agreement. `compare_to_ucsc.py` also computes a fairer
variant (`cnee_vs_nocds_ucsc`): `bedtools subtract`-ing our own per-chromosome CDS bed
(`05-cnees/phastcons/bed/*.cds.bed`, already produced by `extract_cds_bed_chr` from the same
reference GFF - no new CDS extraction needed) from the UCSC bed before comparing.

Separately, `compare_to_atac.py` checks how often each track (our CE/CNEE, both runs, and
UCSC's raw CE) overlaps an independent ATAC-seq peak set - skin, late anagen
(`GSM4156597_skin.late.anagen.peaks.mm39.bed`, translated the same way), producing
`atac-comparison-flat.tsv`.

```{r load-data}
cmp <- read_tsv(repo_file("analyses", "ucsc-mm39-comparison", "ucsc-comparison-flat.tsv"), show_col_types = FALSE) %>%
  mutate(
    mean_len_ours = bp_ours / n_ours,
    mean_len_ucsc = bp_ucsc / n_ucsc,
    jaccard = as.numeric(jaccard)
  )
```

- **Data source**: `data/hamsters/workflow-tests/test-full-split-aln/` (`num_seqs`) and
  `data/hamsters/workflow-tests/test-full-gc/` (`ns`), the same two real, full 21-chromosome
  runs already compared to each other in `resource-usage-and-defaults.Rmd`.
- **UCSC track**: `mm39-35way.bed`
  (`/n/holylfs05/LABS/informatics/Everyone/support/20251114-raftrey-hamster-wga/data/bed/`),
  4,829,248 elements genome-wide; 4,826,807 fall within our 21 target chromosomes after
  translation.
- 126 rows (21 chromosomes x 2 runs x 3 tracks - `ce`, `cnee`, `cnee_vs_nocds_ucsc`) in
  `ucsc-comparison-flat.tsv`; 105 rows (21 chromosomes x 5 sources) in
  `atac-comparison-flat.tsv`.

## Overlap: CE and CNEE vs. UCSC, num_seqs vs. ns

```{r overlap-summary-table}
cmp %>%
  filter(our_track %in% c("ce", "cnee")) %>%
  group_by(our_run, our_track) %>%
  summarise(
    n_chroms = n(),
    total_bp_ours = sum(bp_ours),
    total_bp_ucsc = sum(bp_ucsc),
    total_bp_intersect = sum(bp_intersect),
    mean_jaccard = round(mean(jaccard), 3),
    pct_ours_in_ucsc = round(100 * sum(bp_intersect) / sum(bp_ours), 1),
    pct_ucsc_in_ours = round(100 * sum(bp_intersect) / sum(bp_ucsc), 1),
    .groups = "drop"
  ) %>%
  arrange(our_track, our_run) %>%
  kable(caption = "Aggregate overlap vs. UCSC mm39 35-way, summed across all 21 chromosomes")
```

```{r jaccard-plot, fig.height=6}
ggplot(cmp %>% filter(our_track %in% c("ce", "cnee")), aes(x = reorder(chrom, jaccard), y = jaccard, fill = our_run)) +
  geom_col(position = "dodge") +
  facet_wrap(~our_track, ncol = 1, labeller = labeller(our_track = c(ce = "CEs", cnee = "CNEEs"))) +
  coord_flip() +
  labs(x = NULL, y = "Jaccard vs. UCSC mm39 35-way", fill = "Our run",
       title = "Per-chromosome overlap with the independent UCSC track") +
  theme_minimal()
```

Both our CE and CNEE sets overlap the independent UCSC track substantially above zero
across every chromosome, and the pattern is consistent with what's already known about
this alignment: `CM001014.3` (the poorly cross-species-aligned xy scaffold flagged earlier
this session) shows dramatically lower agreement (Jaccard ~0.04-0.06) than every other
chromosome (~0.2-0.44) - the independent UCSC track corroborates that this specific
chromosome's alignment quality problem is real, not an artifact of our own pipeline.

CE overlap is consistently a bit higher than CNEE overlap for both runs - expected, since
UCSC's track (like our raw CEs) includes coding regions, while our CNEEs have already had
CDS subtracted out. The next section quantifies exactly how much of that gap is just the
CDS-inclusion difference versus a real disagreement.

## A fairer CNEE comparison: subtracting CDS from UCSC's track too

```{r cnee-fair-table}
cmp %>%
  filter(our_track %in% c("cnee", "cnee_vs_nocds_ucsc")) %>%
  group_by(our_run, our_track) %>%
  summarise(
    mean_jaccard = round(mean(jaccard), 3),
    pct_ours_in_ucsc = round(100 * sum(bp_intersect) / sum(bp_ours), 1),
    pct_ucsc_in_ours = round(100 * sum(bp_intersect) / sum(bp_ucsc), 1),
    .groups = "drop"
  ) %>%
  arrange(our_run, our_track) %>%
  kable(caption = "CNEEs vs. UCSC's raw track vs. UCSC with our own CDS bed subtracted first")
```

```{r cnee-fair-plot, fig.height=6}
ggplot(
  cmp %>% filter(our_track %in% c("cnee", "cnee_vs_nocds_ucsc")),
  aes(x = reorder(chrom, jaccard), y = jaccard, fill = our_track)
) +
  geom_col(position = "dodge") +
  facet_wrap(~our_run, ncol = 1) +
  coord_flip() +
  scale_fill_discrete(labels = c(cnee = "vs. raw UCSC", cnee_vs_nocds_ucsc = "vs. CDS-subtracted UCSC")) +
  labs(x = NULL, y = "Jaccard", fill = NULL,
       title = "CNEE overlap: raw UCSC track vs. CDS-subtracted UCSC track, per chromosome") +
  theme_minimal()
```

Subtracting CDS from UCSC's track before comparing raises agreement for both runs (mean
Jaccard: `num_seqs` 0.270 -> 0.310, `ns` 0.259 -> 0.297) - confirming the CE/CNEE gap above
is partly, not entirely, a CDS-inclusion artifact rather than a real difference in what the
two pipelines find non-coding-conserved.

## num_seqs vs. ns: which agrees more with the independent track?

```{r strategy-compare-table}
cmp %>%
  filter(our_track %in% c("ce", "cnee")) %>%
  select(our_run, our_track, chrom, jaccard, pct_ours_in_ucsc, pct_ucsc_in_ours) %>%
  pivot_wider(names_from = our_run, values_from = c(jaccard, pct_ours_in_ucsc, pct_ucsc_in_ours)) %>%
  mutate(jaccard_diff = round(jaccard_num_seqs - jaccard_ns, 4)) %>%
  arrange(our_track, chrom) %>%
  kable(caption = "num_seqs vs. ns agreement with UCSC, per chromosome (positive jaccard_diff = num_seqs agrees more)")
```

## Scale: element counts and lengths

```{r scale-table}
cmp %>%
  filter(our_track %in% c("ce", "cnee")) %>%
  group_by(our_run, our_track) %>%
  summarise(
    total_n_ours = sum(n_ours),
    mean_len_ours_bp = round(sum(bp_ours) / sum(n_ours), 1),
    .groups = "drop"
  ) %>%
  bind_rows(
    cmp %>% filter(our_run == "num_seqs", our_track == "ce") %>%
      summarise(our_run = "-", our_track = "ucsc (per-chrom total)",
                total_n_ours = sum(n_ucsc), mean_len_ours_bp = round(sum(bp_ucsc) / sum(n_ucsc), 1))
  ) %>%
  kable(caption = "Element counts and mean length: ours (both strategies, both tracks) vs. UCSC")
```

UCSC's 35-way track calls roughly 2-3x more elements than either of our runs on the same
chromosomes, consistent with it being built from a much deeper (35 vs 15 species)
alignment - more species generally means more statistical power to detect shorter/weaker
conservation signals. This scale difference is the main reason absolute overlap
percentages sit around 25-40% rather than higher: a majority of UCSC's smaller/weaker
elements simply aren't things a shallower alignment could ever call in the first place,
independent of any real disagreement.

## Overlap with an independent ATAC-seq peak set

How often do CEs/CNEEs from either UCSC or our own pipeline land in independently
accessible chromatin - skin, late anagen ATAC-seq peaks (`GSM4156597_skin.late.anagen`)?

```{r load-atac-data}
atac <- read_tsv(repo_file("analyses", "ucsc-mm39-comparison", "atac-comparison-flat.tsv"), show_col_types = FALSE)
```

```{r atac-summary-table}
atac %>%
  group_by(source) %>%
  summarise(
    total_n_elements = sum(n_elements),
    total_n_overlapping = sum(n_elements_overlapping),
    pct_elements_overlapping = round(100 * sum(n_elements_overlapping) / sum(n_elements), 2),
    mean_jaccard = round(mean(jaccard), 4),
    pct_elements_bp_in_atac = round(100 * sum(bp_intersect) / sum(bp_elements), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_elements_bp_in_atac)) %>%
  kable(caption = "ATAC-seq peak overlap by element count, Jaccard, and bp fraction - summed across all 21 chromosomes")
```

Three different ways of asking "how much overlap" agree on the same ranking (CNEEs >
raw CEs > UCSC), but tell a fuller story together: **by bp fraction**, our CNEEs lead
(~15% of their bp falls in an ATAC peak); **by element count**, an even larger share of
individual CNEEs touch a peak at all (~19% of elements, vs. ~14-16% of CEs) - so it's not
just that CNEEs happen to be bp-heavier where they do overlap, more of them overlap at
all; **Jaccard** (which penalizes for how much of the *union* is peak-only or
element-only space) is lowest for all tracks in absolute terms, as expected given ATAC
peaks and conserved elements are two very different kinds of interval sets in the first
place, but preserves the same relative ranking.

```{r atac-plot, fig.height=6}
ggplot(atac %>% filter(chrom != "CM001014.3"), aes(x = reorder(chrom, pct_elements_bp_in_atac), y = pct_elements_bp_in_atac, fill = source)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "% of track bp overlapping an ATAC-seq peak", fill = "Track",
       title = "Per-chromosome ATAC-seq peak overlap by bp (CM001014.3 excluded - no peaks called)") +
  theme_minimal()
```

```{r atac-jaccard-plot, fig.height=6}
ggplot(atac %>% filter(chrom != "CM001014.3"), aes(x = reorder(chrom, jaccard), y = jaccard, fill = source)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Jaccard vs. ATAC-seq peaks", fill = "Track",
       title = "Per-chromosome ATAC-seq peak overlap by Jaccard (CM001014.3 excluded)") +
  theme_minimal()
```

Our CNEEs overlap ATAC peaks at a modestly *higher* rate than our raw CEs, for both
strategies - consistent with CDS-subtraction concentrating CNEEs toward non-coding,
potentially regulatory sequence rather than diluting them with coding regions that aren't
enriched for this specific tissue's accessible chromatin. UCSC's much larger, shorter-element
track shows the lowest overlap rate of the five, consistent with the "Scale" section above -
a large fraction of its elements are short, weak calls that a shallower alignment (ours) or
a specific regulatory assay (ATAC) wouldn't independently flag either.

## Caveats

- **Not a ground truth.** UCSC's 35-way track is itself a phastCons call on a different
  alignment, not an experimentally validated set of true conserved elements - agreement is
  corroborating evidence, not a correctness proof, and disagreement doesn't necessarily
  mean either pipeline is wrong. The same applies to the ATAC-seq comparison in the other
  direction: it's one tissue, one timepoint - absence of overlap there doesn't mean an
  element isn't real or isn't regulatory in some other context.
- **`mm39-35way.bed`'s `score`/`lod` columns were not used for any filtering** - the full
  track is compared as-is. A confidence-filtered subset might show tighter agreement.
- **The phastCons chrom-name-truncation quirk** (see Methodology) affects only the raw CE
  bed; if any future consumer of `04-phastcons/regions/*.bed` reads it directly by chrom
  name without normalizing, it will silently fail the same way this analysis initially did.
- **`CM001014.3` has zero ATAC-seq peaks** in the source dataset entirely (not specific to
  our pipeline or UCSC) - excluded from the ATAC plot above for that reason, not because of
  any predicted-element issue.
